---
title: C-BARQ-DD (Detector Dog) Descriptive Statistics
author: Liz Hare
date: "`r Sys.Date()`"
email: "lizhare@gmail.com"
output:
    html_document:
        toc: true
        toc_depth: 5
        number_sections: true
---

``` {r, setup, include=FALSE, echo=FALSE}
### load packages
library(tidyverse)
library(lubridate)
library(janitor)
library(knitr)
library(kableExtra)
library(gtsummary)
library(openxlsx)
library(psych)
library(correlation)
library(gluedown) ## inline print lists from R

### get file

dds <- read.xlsx("../intFIles/Detection-dog-survey.xlsx", sheet=1)

### remove records with shifted character fields
### list dog IDs for bad records
badDogs <- c(155436, 155545)
### keep all except bad dog
dds <- dds[!dds$Dog.ID %in% badDogs, ]

### change erroneous character cols to numeric
dds$Fear21 <- as.numeric(dds$Fear21)
dds$Excite35 <- as.numeric(dds$Excite35)
```

# Analysis of Demographic Data


``` {r, demog, results="asis", echo=FALSE}
source("demog.R")
```

## Sex and Neuter Status

there are `r nrow(dds)` dogs in this data set with 
`r nrow(dds[dds$sexF == "male", ])` males and
`r nrow(dds[dds$sexF == "female", ])` females. 
A surprising proportion of dogs are unknown for neuter status
(`r nrow(dds[dds$neuteredF == "unknown", ])/nrow(dds)`).

``` {r, sexTab, results="asis", echo=FALSE}
sexTab <- tbl_summary(dds, by=sexF,
                      include=neuteredF,
                      statistic = all_categorical() ~ "{n}, {p}%")
add_overall(sexTab)



```


## Breed

### Breed Groups

Crossed and mixed breeds were assigned missing values, as were rare breeds that I couldn't 
classify.  


``` {r, breedGroupCount, results="asis", echo=FALSE}
breedGroupTab <- tbl_summary(dds, include=breedGroupF,
                             statistic =  all_categorical() ~ "{n}, {p}%")
breedGroupTab
```

### All Breeds

``` {r, breedCount, results="asis", echo=FALSE}
breedTab <- tbl_summary(dds, include=breedF,
                        statistic = all_categorical() ~ "{n}, {p}%")
breedTab

breedTab
```

## Country

Responses came from `r nlevels(dds$countryF)` countries.

``` {r, countryTab, results="asis", echo=FALSE}
countryTab <- tbl_summary(dds, include=countryF,
                          type = everything() ~ "categorical",
                          statistic = all_categorical() ~ "{n}, {p}%")
countryTab
```

## Age, Training and Retirement

``` {r, ageWorkStatus, results="asis", echo=FALSE}

s1 <- psych::describe(dds$ageY)
s1 <- round(s1, 2)
s1K <- kable(s1, caption="Univariate Statistics for Age in Years")
kable_styling(s1K)

ageWorkStatusTab <- tbl_summary(dds, by=workingStatusF,
                                include=ageY,
                                type = everything() ~ "continuous",
                                statistic = all_continuous() ~ "{mean} ({sd}) ({min} - {max})",
                                digits = list(everything() ~ c(2, 2, 2, 2)))
add_overall(ageWorkStatusTab)
                                              
```

## Types of Work

### Dual Purpose


There are `r nrow(dds[dds$duallyTrainedF == "yes", ])` dual purpose dogs
`r round(nrow(dds[dds$duallyTrainedF == "yes", ])/nrow(dds), digits=2)*100`%).

#### What breeds are dually trained?

``` {r, dualTrain, results="asis", echo=FALSE}
### convert to factor
breedDualTab <- tbl_summary(ddsDual, include=breedF,
                            statistic = all_categorical() ~ "{n}, {p}%")
breedDualTab
```

## Odors Detected

``` {r, odorVars, results="asis", echo=FALSE}
odorTab <- tbl_summary(odorVars, 
                       include=odoraccelerant:odorother,
                       type=everything() ~ "dichotomous",
                       statistic = all_categorical() ~ "{n}, {p}%")
odorTab
```

### Odors entered under "Other"

There are too many responses to print here 
(`r nrow(dds[dds$odorother == "yes" & !is.na(dds$odordescription), ])`). 
Lots of them are nosework-related. Let me know if you want the list.

``` {r, otherOdors, results="asis", echo=FALSE}
##kable(dds[dds$odorother == "yes", c("odordescription")],
##      caption = "Odors Entered Under 'Other'")
```

## Certification

``` {r, certTab, results="asis", echo=FALSE}
certTab <- tbl_summary(certificationVars,
##                         include=cecertificatearda:certificateuspda,
                         type=everything() ~ "dichotomous",
                         statistic = all_categorical() ~ "{n}, {p}%")
certTab
```

### Other Certifying organizations

There are too many answers to print here
(`r nrow(dds[dds$certificateother == "yes" & !is.na(dds$certificatedescription), ])`).

``` {r, otherCert, results="asis", echo=FALSE}
## kable(dds[dds$certificateother == "yes" & !is.na(dds$certificatedescription), c("certificatedescription"), ],
##      caption = "Certificates Entered Under 'Other'")
```

## Alert Type

``` {r, alertTab, results="asis", echo=FALSE}
alertTab <- tbl_summary(dds, include=alerttypeF,
                        type = everything() ~ "categorical",
                        statistic = all_categorical() ~ "{n}, {p}%")
alertTab
```

## Reward System

``` {r, rewardTab, results="asis", echo=FALSE}
rewardTab <- tbl_summary(dds, include=rewardsystemF,
                         type = everything() ~ "categorical",
                         statistic = all_categorical() ~ "{n}, {p}%")
rewardTab
```

## Dogs Handled Before Current

``` {r, dogsHandledTab, results="asis", echo=FALSE}
dogsHandledTab <- tbl_summary(dds, include=dogshandledbeforeF,
                              type = everything () ~ "categorical",
                              statistic = all_categorical() ~ "{n}, {p}%")
dogsHandledTab
```

## Dogs' Home

``` {r, dogsHome, results="asis", echo=FALSE}
dogsHomeTab <- tbl_summary(dds, include=dogshomelocationF,
                           type = everything () ~ "categorical",
                           statistic = all_categorical() ~ "{n}, {p}%")
dogsHomeTab
```

## Owner Rating of Scent Work Ability and General Behavior and Temperament

### Scent Work

``` {r, scenntWorkDesc, results="asis", echo=FALSE}

scentDesc <- psych::describe(dds$scentrating)
scentDesc2 <- round(scentDesc, 2)
scentDescK <- kable(scentDesc2,
                    caption = "Owners' Rating of Dogs' Scent Detection Work Ability")
kable_styling(scentDescK)
```

### General Behavior and Temperament

``` {r, behaviorRate, results="asis", echo=FALSE}
behavDesc <- psych::describe(dds$behaviorrating)
behavDesc <- round(behavDesc, 2)
behavDescK <- kable(behavDesc,
                    caption = "Owners' Rating of Dogs' General Behavior and Temperament")
kable_styling(behavDescK)
```

## DNA Sample Availability

There were `r nrow(dds[dds$sampleprovidedF == "yes", ])` owners willing to provide DNA samples 
(`r round(nrow(dds[dds$sampleprovidedF == "yes", ])/nrow(dds)*100)`%).


# Descriptive Statistics and Plots of C-BARQ-DD Items


``` {r, cleanNames, results="asis", echo=FALSE}
ddItems <- select(dds, c(Train01:Agg20, Fear21:Fear31, SepProb32:SepProb34,
                         Excite35:Excite38, AttAtt39:AttAtt42, Play43:Play46,
                         Frus47:Frus50, Misc51:Misc71))
### pull out items
trainItems <- select(dds, Train01:Train08)
### class(dds$othersituationsagg)
### above variable character so not plotted-- all "othersituations* vars
aggItems <- select(dds, Agg09:Agg20)
fearItems <- select(dds, Fear21:Fear31)
## one fear item is character?
##sapply(fearItems, class)
## 21 is character
### table(fearItems$Fear21, useNA="ifany")
sepItems <- select(dds, SepProb32:SepProb34)
excitItems <- select(dds, Excite35:Excite38)
attItems <- select(dds, AttAtt39:AttAtt42)
playItems <- select(dds, Play43:Play46)
frustItems <- select(dds, Frus47:Frus50)
miscItems <- select(dds, Misc51:Misc71)

```




``` {r, describeItems, results="asis", echo=FALSE}
### are they all numeric?
###table(sapply(ddItems, class))
### why are some character?
### sapply(ddItems, class)
### these are places for verbal additons to quantative answers
### drop them
### dim(    ddItems)
ddItems <- select_if(ddItems, is.numeric)
### dim(ddItems)
###tail(names(ddItems))

save(ddItems, file="~/data/dhs/intFiles/ddCBARQ.Rda")

### summary stats
### count missing

countMissing <- sapply(ddItems, function(x) sum(is.na(x)))
means <- sapply(ddItems, mean, na.rm=TRUE)
sds <- sapply(ddItems, sd, na.rm=TRUE)
mins <- sapply(ddItems, min, na.rm=TRUE)
maxes <- sapply(ddItems, max, na.rm=TRUE)
skews <- sapply(ddItems, skew, na.rm=TRUE)
kurts <- sapply(ddItems, kurtosi, na.rm=TRUE)
### shapiro-wilk test for normality
sws <- sapply(ddItems, function(x) shapiro.test(x)$statistic)
swp <- sapply(ddItems, function(y) shapiro.test(y)$p.value)

bigDesc <- data.frame(means, sds, mins, maxes, skews, kurts, sws, swp)
colnames(bigDesc) <- c("Mean", "SD", "Min", "Max", "Skewness", "Kurtosis",
                       "Shapiro-Wilk Score", "Shapiro-Wilk p")
bigDescK <- kable(bigDesc, digits=2,
      caption="Descriptive Statistics for Detector-C-BARQ Items")

kable_styling(bigDescK)

largeSkew <- bigDesc[bigDesc$skews < -2 & bigDesc$skews > 2, ]

largeSkewItems <- rownames(bigDesc[bigDesc$Skewness < -2 & bigDesc$skews > 2, ])
largeKurtItems <- rownames(bigDesc[bigDesc$Kurtosis > 7, ])
### all items you'd remove
removeSK <- unique(c(largeSkewItems, largeKurtItems))
### new df without non-normal items
n1 <- ddItems[ , -which(colnames(ddItems) %in% removeSK)]
dim(n1)

```

According to Watkins (2021), items should be removed if they have skewness > 2.0 or 
kurtosis > 7.0. These items would violate the assumptions of the Pearson correlation coefficients 
that factor analysis is based on. (Do we need to consider the items as ordinal discrete values?)  

Items with skewness > 2.0 include 
`r md_bullet(largeSkewItems)`

items with kurtosis > 7.0 include
	`r md_bullet(largeKurtItems)`

For the current data set, `r length(removeSK)`
items would be removed for the factor analysis, leaving
	`r ncol(n1)` to analyze.

### Count Plots

#### Trainability


``` {r, plotTrainFreq, results="asis", echo=FALSE, fig.width=12, fig.height=24}
### plot



plotFreq <- function(x) {
    longData <- pivot_longer(x, cols = everything(),
                             names_to = "item", values_to = "score")
    ggplot(longData, aes(x = score)) +
                     geom_bar(na.rm=TRUE) +
                     facet_wrap(vars(item), ncol=4) +
                     labs(x = "Score", y="Count",
                          title="Counts")
    }
plotFreq(trainItems) 
```

#### Aggression

``` {r, plotAggFreq, results="asis", echo=FALSE}
plotFreq(aggItems)
```

#### Fear

``` {r, plotFearFreq, results="asis", echo=FALSE}
plotFreq(fearItems)
```


#### Separation

``` {r, plotSepFreq, results="asis", echo=FALSE}
plotFreq(sepItems)
```


#### Excitability

``` {r, plotExcitFreq, results="asis", echo=FALSE}
plotFreq(excitItems)
```

#### Attention-Seeking

``` {r, plotAttFreq, results="asis", echo=FALSE}
plotFreq(attItems)
```



#### Play

``` {r, plotPlayFreq, results="asis", echo=FALSE}
plotFreq(playItems)
```


#### Frustration

``` {r, plotFrustFreq, results="asis", echo=FALSE}
plotFreq(frustItems)
```


#### Miscellaneous

``` {r, plotMiscFreq, results="asis", echo=FALSE}
plotFreq(miscItems)
```

```
